Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-06-14 doc/thesis/discussion.tex
Writing from last night

@@

Contact networks can have a strong influence on epidemic progression, and are
potentially useful as a public health tool~\autocite{wang2015targeting,
little2014using}. Despite this, few methods exist for investigating contact
network parameters in a phylodynamic framework \autocite[although see][for
related work]{groendyke2011bayesian, volz2008sir, brown2011transmission,
leventhal2012inferring, greenbaum2016inference}. Kernel-ABC is a model-agnostic
method which can be used to investigate any quantity that affects tree
shape~\autocite{poon2015phylodynamic}. In this work, we developed
\software{netabc}, a method based on kernel-ABC to infer the parameters of a
contact network model. The method is general, meaning that it can be used to
infer parameters of any network model, as long as it allows simulated networks
can be easily generated. We have included generators for the \gls{BA} model
discussed here, as well as the \gls{ER} and \gls{WS} network models.
Instructions for adding additional models are available in the project's online
documentation. We have made \software{netabc} publicly available at
\url{github.com/rmcclosk/netabc} under a permissive open source license, to
encourage other researchers to apply and extend our method.

Several alternative network models and modelling frameworks have been developed
which may provide useful future targets for kernel-assisted \gls{ABC}. Waring
models~\autocite{irwin1963place,handcock2004likelihood} are a more flexible
type of preferential attachment model which permit a subset of attachments to
be formed non-preferentially. These models were used by
\textcite{brown2011transmission} to characterize the transmission network in
the United Kingdom. \Glspl{ERGM}~\autocite{robins2007introduction} are a
flexible and expressive parameterization of contact networks in terms of
statistics of network features such as pairs and triads.
\textcite{goodreau2006assessing} evaluated the effect of several different
\gls{ERGM} parameterizations on transmission tree shape and effective
population size. The author suggested the use of \glspl{ERGM} as a general
framework for estimation of epidemiological quantities related to \gls{HIV}
transmission. Except for a few special cases, simulating a network according to
an \gls{ERGM} generally requires \gls{MCMC}, which would be too computationally
intensive to integrate into \software{netabc} as it currently stands. To fit
\gls{ERGM} with kernel-assisted \gls{ABC}, one possibility would be to consider the
network itself as a parameter to be modified by the \gls{MCMC} kernel.
Other network modelling frameworks include the partnership-centric formulation
developed by~\textcite{eames2002modeling} and the log-linear adjacency matrix
parameterization applied by~\textcite{morris1993epidemiology}.

%\Textcite{morris1993epidemiology} proposes to apply the standard compartmental
%modelling framework to contact networks by assigning each individual their own
%compartment. Thus, each individual is associated with a single \gls{ODE}, with
%the entire \gls{ODE} system parameterized by the adjacency matrix of the
%contact network. 
%The author proposes to use log-linear models to parameterize the adjacency
%5matrix of a contact network. This framework is highly expressive, and allows
%straightforward incorporation of time-dependent dynamics. 
%However, simulating a transmission tree would require the numerical solution of
%a very large system of \glspl{ODE}, which may be computationally prohibitive
%given the number of simulations required for kernel-assisted \gls{ABC}.

The two-step process of simulating a contact network and subsequently allowing
an epidemic to spread over that network carries with it the assumption that the
contact network is static over the duration of the epidemic. Clearly this
assumption is invalid, as people make and break partnerships on a regular
basis. Addressing the impact of this simplifying assumption is outside the
scope of this work. However, the same assumption is made by most studies using
contact network models in an epidemiological
context~\autocite{welch2011statistical, bansal2007individual}. In principle,
kernel-assisted \gls{ABC} could be adapted to dynamic contact networks by using
a method such as that developed by \textcite{robinson2012dynamics} to simulate
such a network, while concurrently simulating the spread of an epidemic.

It is important to note that \software{netabc} takes a transmission tree as
input, rather than a viral phylogeny. In reality, true transmission trees are
not available and must be estimated; these estimates are often based on the
viral phylogeny. Although this has been demonstrated to be a fair
approximation~\autocite[e.g.][]{leitner1996accurate}, and is frequently used in
practice~\autocite[e.g.][]{stadler2013uncovering}, the topologies of a viral
phylogeny and transmission tree can differ
significantly~\autocite{ypma2013relating, hall2015epidemic} due to within-host
evolution and the sampling process. We have left the estimation of a
transmission tree up to the user. In theory, it is possible to incorporate the
process by which a viral phylogeny is generated along with a transmission tree
into our method, for example by simulating within-host dynamics. {\color{blue}
\uline{Indeed, progress has very recently been demonstrated on this front in a
talk by \citeauthor{giardina2016inference}, who have independently developed in
a method similar to ours to fit contact network models to phylogenetic data
that additionally incorporates a within-host evolutionary model.}} Although
this may be an avenue for future extension, we felt that it would obscure the
primary purpose of this work, which is to study contact network parameters. In
addition, there are a number of different methods available for inferring
transmission trees~\autocite{didelot2014bayesian, ypma2012unravelling,
jombart2011reconstructing, cottam2008integrating, hall2015epidemic}, some of
which incorporate geographic and/or epidemiological data not accommodated by
our method. We therefore felt it would be best to allow researchers to use
their own preferred method of constructing a transmission tree.

@@

The preferential attachment power \gls{alpha} had a very strong influence on
tree shape in the range of values we considered
(\cref{fig:alphatrees,fig:kpca}). Although the tree kernel was the most
effective classifier for \gls{alpha}, a Sackin's index of tree imbalance
performed nearly as well (\cref{fig:rsquared}). This result was intuitive: high
\gls{alpha} values produce networks with few well-connected ``superspreader''
nodes which are involved in a large number of transmissions, resulting in a
highly unbalanced ladder-like tree structure (\cref{fig:alphatrees}). 
There was no observable bias in the estimates of \gls{alpha} obtained with
\software{netabc}, however {\color{red}\sout{the variation in these estimates
was higher for $\alpha < 1$ than for $\alpha \geq 1$}} {\color{blue}\uline{there
appeared to be weaker identifiability for \gls{alpha} $<$ 1 than for \gls{alpha}
$\geq$ 1}} (\cref{fig:abcptm2,tab:abchpd}). The relationship between
\gls{alpha} and the power law exponent \gls{gamma} may explain this result
(\cref{fig:gamma}). The \gls{gamma} values associated with \gls{alpha} = 0 and
\gls{alpha} = 0.5 are nearly identical (about 2.28 for \gls{alpha} = 0 and 2.33
for \gls{alpha} = 0.5 with \gls{N} = 5000 and \gls{m} = 2). In other words, the
degree distributions of networks with \gls{alpha} $< 1$ are similar to each
other, which may result in similarity of corresponding transmission trees as
well. 
%The accuracy of \gls{alpha} estimates also increased when the value of the
%\gls{m} parameter was higher (compare
%\cref{fig:abcptm2,fig:abcptm3,fig:abcptm4}).

The \gls{I} parameter, representing the prevalence at the time of sampling, was
also generally identifiable, although it was slightly over-estimated for both
cases we considered with kernel-assisted \gls{ABC}. The dynamics of the \gls{SI} model,
and the coalescent process~\autocite{kingman1982coalescent}, offer a potential
explanation for the identifiability of \gls{I}. In our simulations, we assumed
that all discordant edges shared the same transmission rate, so that the
waiting time until the next transmission in the entire network was always
inversely proportional to the number of discordant edges. In the initial phase
of the epidemic, when \gls{I} is small, each new transmission results in many
new discordant edges. Hence, there is an early exponential growth phase,
producing many short branches near the root of the tree. As the epidemic gets
closer to saturating the network, the number of discordant edges decays,
causing longer waiting times. The distribution of coalescence times in the tree
should therefore be informative about \gls{I}~\autocite{volz2009phylodynamics}.
This information is captured by the tree kernel, and also by the \gls{nltt}
statistic, which both performed quite well in classifying \gls{I}
(\cref{fig:rsquared}).

The number of nodes in the network, \gls{N}, exhibited the most variation in
terms of its effect on tree shape. There was almost no difference between trees
simulated under different \gls{N} values when the number of infected nodes
\gls{I} was small. There is an intuitive explanation for this result, namely
that adding additional nodes does not change the edge density or overall shape
of a \gls{BA} network. This can be illustrated by imagining that we add a small
number of nodes to a network after the epidemic simulation has already been
completed. It is possible that none of these new nodes attains a connection to
any infected node. Thus, running the simulation again on the new, larger
network could produce the exact same transmission tree as before. On the other
hand, when \gls{I} is large relative to \gls{N}, the coalescent dynamics
discussed above also apply. That is, the waiting times until the next infection
increase, resulting in longer coalescence times toward the tips. The relative
accuracy of the \gls{nltt} in these situations
(\cref{fig:rsquared,fig:Ncrossv}) corroborates this hypothesis, as the
\gls{nltt} uses only information about the coalescence times. When all \gls{BA}
parameters were simultaneously estimated with kernel-assisted \gls{ABC}, \gls{N} was
nearly always over-estimated by approximately a factor of two
(\cref{fig:abcptm2,tab:abchpd}). One factor which may have contributed to this
bias was our choice of prior distribution. Since the prior for \gls{I} and
\gls{N} was jointly uniform on a region where $I \leq N$, more prior weight was
assigned to higher \gls{N} values. Another contributing factor relates to the
dynamics of the \gls{SI} model and the coalescent process.

\gls{I} and \gls{N} were both systematically over-estimated by
\software{netabc}, although the bias was more severe for \gls{N} than for
\gls{I}. The number of infected individuals follows a logistic growth curve
under the \gls{SI} model. This kind of growth curve has three qualitative
phases: a slow ramp-up, an exponential growth phase, and a slow final phase
when the susceptible population is almost depleted. The waiting times until the
next transmission, which determine the coalescence times in the tree, are
dependent on the growth phase of the epidemic. Therefore, we hypothesize that
it is the growth phase at the time of sampling which most affects tree shape,
rather than the specific values of \gls{I} or \gls{N}. To investigate this
hypothesis, we simulated transmission trees over networks on a grid of \gls{I}
and \gls{N} values in the region of uniform prior density. We fit logistic
growth curves to the proportion of infected individuals over time, and
calculated the first and second derivatives of these curves at the time of
transmission tree sampling. These derivatives give us an indication of the
growth rates of the epidemics at the time of sampling. As shown in
\cref{fig:derivs}, there are bands along which both derivatives are similar
which contain the values we tested. These bands span mostly higher values of
\gls{N} and \gls{I} than the true values. Therefore, if \gls{N} and \gls{I} are
free to vary (as is the case in kernel-assisted \gls{ABC}), and our hypothesis is true,
both parameters will tend to be overestimated due to being less identifiable
within their own band. However, when \gls{N} is fixed at 5000, the derivatives
vary substantially along the \gls{I}-axis, which explains why the grid search
estimates of \gls{I} were accurate and unbiased (\cref{fig:gridptI}).

\begin{figure}[ht]
  \centering
  \includegraphics{derivatives.pdf}
  \caption[
    First and second derivatives of epidemic growth curves at time of sampling
    for various values of $I$ and $N$.
  ]{
    First and second derivatives of epidemic growth curves at time of sampling
    for various values of \gls{I} and \gls{N}. Networks were simulated under
    the \gls{BA} model with \gls{alpha} = 1.0, \gls{m} = 2, and \gls{N} varied 
    along the values shown on the $x$-axis. Transmission trees were sampled at
    the time when \gls{I} nodes were infected ($y$-axis). Logistic growth
    curves were fit to epidemic trajectories derived from the transmission
    trees, and their first and second derivatives were calculated at the time
    of sampling. Contours show first derivatives, while colours indicate second
    derivatives. Values of \gls{I} and \gls{N} used in simulation experiments
    with kernel-assisted \gls{ABC} are indicated by diamonds.
  }
  \label{fig:derivs}
\end{figure}

The \gls{m} parameter, which controls the number of connections added to the
network per vertex, did not have a measurable impact on tree shape and was not
identifiable with kernel-ABC. The exception to this was the value \gls{m} = 1,
which produces networks without cycles whose associated trees were more easily
distinguished. However, all the analyses presented here did not take the
absolute size of the transmission trees into account, as the branch lengths
were rescaled by their mean. Because higher \gls{m} values imply higher edge
density, an epidemic should spread more quickly for higher \gls{m} than lower
\gls{m} with the same per-edge transmission probability.  Hence, considering
the absolute height of the trees may improve our method's ability to
reconstruct \gls{m}.

@@

@@

For all datasets we examined, the posterior mean estimates for \gls{alpha} were
sub-linear, ranging from 
    0.27
to 
    0.95.
The sub-linearity is consistent with the results of
\textcite{de2007preferential}, who developed a statistical inference method to
estimate the parameters of a more sophisticated \acrlong{PA} model
incorporating heterogeneous node behaviour. They found \gls{alpha} values
ranging from 0.26 to 0.62, depending on the gender and time period considered. 
Our estimates of \gls{alpha} for the \textcite{niculescu2015recent} was above
this range under both priors, as were the estimates for the
\textcite{wang2015targeting} data and the BC data when \gls{m} = 1 was
disallowed by the prior. The dataset investigated by
\textcite{de2007preferential} was derived from a survey of a random sample of
the Norwegian population, whereas our investigation focused on datasets from
known phylogenetic or geographic clusters of \gls{HIV} infected persons. It is
therefore unsurprising that we detected stronger \acrlong{PA} dynamics in some
cases. For instance, random sampling is much less likely to discover the
high-degree nodes characterizing the tail of the degree distribution, simply
because those individuals are rare in the general population. In addition, it
is plausible that \gls{HIV}-positive individuals are more likely to be highly
connected in their sexual networks, as the odds of acquiring \gls{HIV} increase
with the number of unprotected sexual contacts.

Both \textcite{de2007preferential} and \textcite{novitsky2014impact} studied
populations whose primary risk factor for \gls{HIV} infection was heterosexual
contact. \citeauthor{de2007preferential} explicitly excluded reported
homosexual contacts; \citeauthor{novitsky2014impact} did not, but noted that
heterosexual contact is the primary mode of transmission in Botswana where the
study was done. In the first of the two papers describing the Botswana
study~\autocite{novitsky2013phylogenetic}, the authors noted that their sample
was gender-biased, being composed of approximately 75\% women. Our estimate of
\gls{alpha} for these data was 
    0.55, 0.55
        or 
    0.53, 0.53,
depending on the prior on \gls{m}; \citeauthor{de2007preferential} estimated
0.54, 0.57, and 0.29 for 3-year, 5-year, and lifetime partnership networks
respectively for the female portion of their sample.

>>>>

%<<setup, include=FALSE>>=
%    source("global.R")
%@

@@

\glsreset{ER}
\glsreset{WS}

Contact networks can have a strong influence on epidemic progression, and are
potentially useful as a public health tool~\autocite{wang2015targeting,
little2014using}. Despite this, few methods exist for investigating contact
network parameters in a phylodynamic framework \autocite[although see][for
related work]{groendyke2011bayesian, volz2008sir, brown2011transmission,
leventhal2012inferring, greenbaum2016inference}. Kernel-ABC is a model-agnostic
method which can be used to investigate any quantity that affects tree
shape~\autocite{poon2015phylodynamic}. In this work, we developed
\software{netabc}, a method based on kernel-ABC to infer the parameters of a
contact network model. The method \del{is general, meaning that it} can be used
to infer parameters of any network model, as long as it allows simulated
networks to be easily generated. We have included generators for the \gls{BA}
model discussed here, as well as the \gls{ER}~\autocite{erdos1960evolution} and
\gls{WS}~\autocite{watts1998collective} network models.  Instructions for
adding additional models are available in the project's online documentation.
We have made \software{netabc} publicly available at
\url{github.com/rmcclosk/netabc} under a permissive open source license, to
encourage other researchers to apply and extend our method.

Several alternative network models and modelling frameworks have been developed
which may provide useful future targets for \gls{ABC}. Waring
models~\autocite{irwin1963place,handcock2004likelihood} are a more flexible
type of preferential attachment model which permit a subset of attachments to
be formed non-preferentially. These models were used by
\textcite{brown2011transmission} to characterize the transmission network in
the United Kingdom. \Glspl{ERGM}~\autocite{robins2007introduction} are a
flexible and expressive parameterization of contact networks in terms of
statistics of network features such as pairs and triads.
\textcite{goodreau2006assessing} evaluated the effect of several different
\gls{ERGM} parameterizations on transmission tree shape and effective
population size, and suggested the use of \glspl{ERGM} as a general framework
for estimation of epidemiological quantities related to \gls{HIV} transmission.
Except for a few special cases, simulating a network according to an \gls{ERGM}
generally requires \gls{MCMC}, which would be too computationally intensive to
integrate into \software{netabc} as it currently stands. To fit \gls{ERGM} with
\gls{ABC}, one possibility would be to consider the network
itself as a parameter to be modified by the \gls{MCMC} kernel.  Other network
modelling frameworks include the partnership-centric formulation developed
by~\textcite{eames2002modeling} and the log-linear adjacency matrix
parameterization applied by~\textcite{morris1993epidemiology}.

The two-step process of simulating a contact network and subsequently allowing
an epidemic to spread over that network carries with it the assumption that the
contact network is static over the duration of the epidemic. Clearly this
assumption is false, as people make and break partnerships on a regular basis.
Addressing the impact of this simplifying assumption is outside the scope of
this work. However, the same assumption is made by most studies using contact
network models in an epidemiological context~\autocite{welch2011statistical,
bansal2007individual}. In principle, \gls{ABC} could be adapted
to dynamic contact networks by using a method such as that developed by
\textcite{robinson2012dynamics} to simulate such a network, while concurrently
simulating the spread of an epidemic.

It is important to note that \software{netabc} takes a transmission tree as
input, rather than a viral phylogeny. In reality, true transmission trees are
not available and must be estimated; these estimates are often based on the
viral phylogeny. Although this has been demonstrated to be a fair
approximation~\autocite[e.g.][]{leitner1996accurate}, and is frequently used in
practice~\autocite[e.g.][]{stadler2013uncovering}, the topologies of a viral
phylogeny and transmission tree can differ
significantly~\autocite{ypma2013relating, hall2015epidemic} due to within-host
evolution and the sampling process. We have left the estimation of a
transmission tree up to the user. In theory, it is possible to incorporate the
process by which a viral phylogeny is generated along with a transmission tree
into our method, for example by simulating within-host dynamics. \add{Indeed,
progress has very recently been demonstrated on this front in a talk by
\textcite{giardina2016inference}, who have independently developed in a method
similar to ours to fit contact network models to phylogenetic data that
additionally incorporates a within-host evolutionary model.} Although this may
be an avenue for future extension, we felt that it would obscure the primary
purpose of this work, which is to study contact network parameters. In
addition, there are a number of different methods available for inferring
transmission trees~\autocite{didelot2014bayesian, ypma2012unravelling,
jombart2011reconstructing, cottam2008integrating, hall2015epidemic}, some of
which incorporate geographic and/or epidemiological data not accommodated by
our method. We therefore felt it would be best to allow researchers to use
their own preferred method of constructing a transmission tree.

@@

The preferential attachment power \gls{alpha} had a very strong influence on
tree shape in the range of values we considered
(\cref{fig:alphatrees,fig:kpca}). Although the tree kernel was the most
effective classifier for \gls{alpha}, a Sackin's index of tree imbalance
performed nearly as well (\cref{fig:rsquared}). This result was intuitive: high
\gls{alpha} values produce networks with few well-connected ``superspreader''
nodes which are involved in a large number of transmissions, resulting in a
highly unbalanced ladder-like tree structure (\cref{fig:alphatrees}). 
\del{There was no observable bias in the estimates of \gls{alpha} obtained with
\software{netabc}, however the variation in these estimates
was higher for $\alpha < 1$ than for $\alpha \geq 1$} \add{There appeared to be
weaker identifiability for \gls{alpha} $<$ 1 than for \gls{alpha} $\geq$ 1}
(\cref{fig:abcptm2,tab:abchpd}), which may be partially explained by the
relationship between \gls{alpha} and the power law exponent \gls{gamma}
(\cref{fig:gamma}). \add{Although the degree distributions do not truly follow
a power law for $\gls{alpha} \neq 1$, the fitted exponent still captures the
slope of the degree distribution reasonably well (\cref{fig:powerlaw}), and 
distributions with similar fitted \gls{gamma} values appear qualitatively
similar in other respects.} The \gls{gamma} values fitted to \gls{alpha} = 0
and \gls{alpha} = 0.5 are nearly identical (about 2.28 for \gls{alpha} = 0 and
2.33 for \gls{alpha} = 0.5 with \gls{N} = 5000 and \gls{m} = 2). In other
words, the degree distributions of networks with \gls{alpha} $< 1$ are similar
to each other, which may result in similarity of corresponding transmission
trees as well.

The \gls{I} parameter, representing the prevalence at the time of sampling, was
also generally identifiable, although it was slightly over-estimated for both
cases we considered with \gls{ABC}. \add{Sackin's index was better able to
capture the effect of \gls{I} on tree shape than the \gls{nltt}
(\cref{fig:rsquared,fig:Icrossv}), suggesting that this parameter impacts the
distribution of branching times in the tree more than the topology. In a
homogeneously-mixed population, branching times can be modelled by the
coalescent process~\autocite{kingman1982coalescent}, in our case under the
\gls{SI} model~\autocite{volz2009phylodynamics}. Although our populations are
not homogeneously mixed, the forces which affect the distributions of branching
times still apply.} \del{The dynamics of the \gls{SI} model, and the coalescent
process, offer a potential explanation for the identifiability of \gls{I}.} In
our simulations, we assumed that all discordant edges shared the same
transmission rate, so that the waiting time until the next transmission in the
entire network was always inversely proportional to the number of discordant
edges. In the initial phase of the epidemic, when \gls{I} is small, each new
transmission results in many new discordant edges. Hence, there is an early
exponential growth phase, producing many short branches near the root of the
tree. As the epidemic gets closer to saturating the network, the number of
discordant edges decays, causing longer waiting times. \del{The distribution of
coalescence times in the tree should therefore be informative about \gls{I}.
This information is captured by the tree kernel, and also by the \gls{nltt}
statistic, which both performed quite well in classifying \gls{I}
(\cref{fig:rsquared}).}

The number of nodes in the network, \gls{N}, exhibited the most variation in
terms of its effect on tree shape. There was almost no difference between trees
simulated under different \gls{N} values when the number of infected nodes
\gls{I} was small (\cref{fig:Nkpca,fig:Ncrossv}). There is an intuitive
explanation for this result, namely that adding additional nodes does not
change the edge density or overall shape of a \gls{BA} network. \add{In fact,
this is why the degree distributions of these networks are called
``scale-free.''} This can be illustrated by imagining that we add a small
number of nodes to a network after the epidemic simulation has already been
completed. It is possible that none of these new nodes attains a connection to
any infected node. Thus, running the simulation again on the new, larger
network could produce the exact same transmission tree as before. On the other
hand, when \gls{I} is large relative to \gls{N}, the coalescent dynamics
discussed above also apply. The waiting times until the next infection
increase, resulting in longer coalescence times toward the tips. The relative
accuracy of the \gls{nltt} in these situations
(\cref{fig:rsquared,fig:Ncrossv}) corroborates this hypothesis, as the
\gls{nltt} uses only information about the coalescence times. When all \gls{BA}
parameters were simultaneously estimated with \gls{ABC}, \gls{N} was nearly
always over-estimated by approximately a factor of two
(\cref{fig:abcptm2,tab:abchpd}). One factor which may have contributed to this
bias was our choice of prior distribution. Since the prior for \gls{I} and
\gls{N} was jointly uniform on a region where $I \leq N$, more prior weight was
assigned to higher \gls{N} values. Another contributing factor relates to the
dynamics of the \gls{SI} model and the coalescent process and is discussed
below.

\gls{I} and \gls{N} were both systematically over-estimated by
\software{netabc}, although the bias was more severe for \gls{N} than for
\gls{I}. The number of infected individuals follows a logistic growth curve
under the \gls{SI} model. This kind of growth curve has three qualitative
phases: a slow ramp-up, an exponential growth phase, and a slow final phase
when the susceptible population is almost depleted. The waiting times until the
next transmission, which determine the coalescence times in the tree, are
dependent on the growth phase of the epidemic. Therefore, we hypothesize that
it is the growth phase at the time of sampling which most affects tree shape,
rather than the specific values of \gls{I} or \gls{N}. To investigate this
hypothesis, we simulated transmission trees over networks on a grid of \gls{I}
and \gls{N} values in the region of uniform prior density. We fit logistic
growth curves to the proportion of infected individuals over time, and
calculated the first and second derivatives of these curves (with respect to
time) at the time of transmission tree sampling. These derivatives give us an
indication of the growth rates of the epidemics at the time of sampling. As
shown in \cref{fig:derivs}, there are bands along which both derivatives are
similar which contain the values we tested. These bands span mostly higher
values of \gls{N} and \gls{I} than the true values. Therefore, if \gls{N} and
\gls{I} are free to vary (as is the case in \gls{ABC}), and our hypothesis is
true, both parameters will tend to be overestimated due to being less
identifiable within their own band. However, when \gls{N} is fixed at 5000, the
derivatives vary substantially along the \gls{I}-axis, which explains why the
grid search estimates of \gls{I} were accurate and unbiased
(\cref{fig:gridptI}).

\begin{figure}[ht]
  \centering
  %\includegraphics{derivatives.pdf}
  \caption[
    First and second time derivatives of epidemic growth curves at time of
    sampling for various values of $I$ and $N$.
  ]{
    First and second \add{time} derivatives of epidemic growth curves at time
    of sampling for various values of \gls{I} and \gls{N}. Networks were
    simulated under the \gls{BA} model with \gls{alpha} = 1.0, \gls{m} = 2, and
    \gls{N} varied along the values shown on the $x$-axis. Transmission trees
    were sampled at the time when \gls{I} nodes were infected ($y$-axis).
    Logistic growth curves were fit to epidemic trajectories derived from the
    transmission trees, and their first and second derivatives were calculated
    at the time of sampling. Contours show first derivatives, while colours
    indicate second derivatives. Values of \gls{I} and \gls{N} used in
    simulation experiments with \gls{ABC} are indicated by diamonds.
  }
  \label{fig:derivs}
\end{figure}

The \gls{m} parameter, which controls the number of connections added to the
network per vertex, did not have a measurable impact on tree shape and was not
identifiable with \gls{ABC}, \add{nor with grid search when the other
parameters were held fixed.} The exception to this was the value \gls{m} = 1,
which produces networks without cycles whose associated trees were more easily
distinguished (\cref{fig:gridest,fig:gridm}). However, all the analyses
presented here did not take the absolute size of the transmission trees into
account, as the branch lengths were rescaled by their mean. Because higher
\gls{m} values imply higher edge density, an epidemic should spread more
quickly for higher \gls{m} than lower \gls{m} with the same per-edge
transmission probability.  Hence, considering the absolute height of the trees
may improve our method's ability to reconstruct \gls{m}. \add{It was also
pointed out to us by an anonymous reviewer that for a fixed \gls{I}, an
infected node may only end up transmitting along a fraction of its outgoing
edges, which could mask the presence of the extra edges associated with higher
\gls{m}}.

\end{document}

@@

%<<disc, include=FALSE>>=
%    d <- setDT(collect.data("../../simulations/aggregates/hpd/*"))
%@

@@

For all datasets we examined, the posterior mean estimates for \gls{alpha} were
sub-linear, ranging from 
    X%\S%expr{d[m_min == 1 & parameter == "alpha", round(min(mean), 2)]}
to 
    X%\S%expr{d[m_min == 1 & parameter == "alpha", round(max(mean), 2)]}.
The sub-linearity is consistent with the results of
\textcite{de2007preferential}, who developed a statistical inference method to
estimate the parameters of a more sophisticated \acrlong{PA} model
incorporating heterogeneous node behaviour. They found \gls{alpha} values
ranging from 0.26 to 0.62, depending on the gender and time period considered. 
Our estimates of \gls{alpha} for the \textcite{niculescu2015recent} was above
this range under both priors, as were the estimates for the
\textcite{wang2015targeting} data and the BC data when \gls{m} = 1 was
disallowed by the prior. The dataset investigated by
\textcite{de2007preferential} was derived from a survey of a random sample of
the Norwegian population, whereas our investigation focused on datasets from
known phylogenetic or geographic clusters of \gls{HIV} infected persons. It is
therefore unsurprising that we detected stronger \acrlong{PA} dynamics in some
cases. For instance, random sampling is much less likely to discover the
high-degree nodes characterizing the tail of the degree distribution, simply
because those individuals are rare in the general population. In addition, it
is plausible that \gls{HIV}-positive individuals are more likely to be highly
connected in their sexual networks, as the odds of acquiring \gls{HIV} increase
with the number of unprotected sexual contacts.

Both \textcite{de2007preferential} and \textcite{novitsky2014impact} studied
populations whose primary risk factor for \gls{HIV} infection was heterosexual
contact. \citeauthor{de2007preferential} explicitly excluded reported
homosexual contacts; \citeauthor{novitsky2014impact} did not, but noted that
heterosexual contact is the primary mode of transmission in Botswana where the
study was done. In the first of the two papers describing the Botswana
study~\autocite{novitsky2013phylogenetic}, the authors noted that their sample
was gender-biased, being composed of approximately 75\% women. Our estimate of
\gls{alpha} for these data was 
    X%\S%expr{d[m_min == 1 & dataset == "novitsky2014" & parameter == "alpha", round(mean, 2)]}
        or 
    X%\S%expr{d[m_min == 2 & dataset == "novitsky2014" & parameter == "alpha", round(mean, 2)]},
depending on the prior on \gls{m}; \citeauthor{de2007preferential} estimated
0.54, 0.57, and 0.29 for 3-year, 5-year, and lifetime partnership networks
respectively for the female portion of their sample.
